Machine learning high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-12-28 paper1.tex
acknowledgements

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Emulation of tree-level $e^{+}e^{-}$ matrix elements}
\label{chapter:fame1}

\section{Introduction}
\label{sec:intro}
The capacity of neural networks to approximate intricate functions have already been 
used to provide fast calculations of production cross-sections~\cite{Otten:2018kum,Buckley:2020bxg}.
In this chapter we investigate whether NNs can approximate production cross-sections more differentially by replacing computationally expensive 
matrix element calculations. 
The challenge with this endeavour is that matrix elements are plagued with numerous divergences that arise from infrared divergences.
In previous works \cite{Badger:2020uow,Aylett-Bullock:2021hmo} a combination of individual neural networks were used 
to approximate matrix elements. In order to deal with the complex structure of the matrix elements 
the authors of these studies divided the phase-space into sectors according to the infrared singularities and trained networks on these sectors,
thereby limiting the complexity of the fit by isolating a single divergence per sector. All sectors were then combined to make the final prediction.
While the authors of the initial study for electron-positron annihilation showed good agreement between the total cross-section and histogrammed distributions both at LO and NLO,
we note that the accuracy of the interpolation at the level of individual points was a lot worse than when averaged in the 
histograms. In addition, the performance of the extrapolation outside of its training phase-space 
(i.e. more singular configurations than those considered to fit the model) is problematic.

@@

This chapter is organised as follows. 
Section~\ref{sec:fitting} introduces our factorisation-aware 
deep neural network model, Section~\ref{sec:results} showcases our results with a comparison with the 
model from Ref.~\cite{Badger:2020uow} in Section~\ref{sec:n3jet_comparison}, and an analysis of the performance of our model in Section~\ref{sec:mainresult}. 
To show that the neural network is both interpolating and extrapolating well we show its 
behaviour on random trajectories in phase-space in Section~\ref{sec:trajectories}. 
Our conclusions are presented in Section~\ref{sec:conclusion} 
and some discussion of specific details are collected in the appendices so as not to distract from the main discussion.

@@

A neural network can be seen as a function $f(\vec{x}; \theta) = \vec{y}$, where $f: \mathbb{R}^{d} \rightarrow \mathbb{R}$ maps a 
$d$-dimensional vector $\vec{x}$ of inputs onto the vector of outputs $\vec{y}$, and where $\theta$ are the parameters of the neural network which we aim to optimise 
such that the outputs $\vec{y}$ of the neural network match the target as well as possible. 
The simplest implementation of an emulator would be to take the input as the kinematic information of the phase-space point and the output to be the full matrix element. 
Our approach modifies both the input of the NN and its target, as described in the following sections.

\subsection{Infrared divergences and dipole factorisation formula}
It is well known that in soft and collinear limits the matrix element in $(n+1)$-body phase-space factorises into a singular factor and a reduced matrix element in $n$-body phase-space \cite{Altarelli:1977zs,BASSETTO1983201} (Section~\ref{sec:me_factorisation}).
This factorisation was used by Catani and Seymour \cite{Catani:1996vz} to construct subtraction terms for the real radiation part of a NLO calculation (Section~\ref{sec:CS_dipoles}).
They introduced a factorisation formula with universal dipoles that smoothly interpolates 
between the soft and collinear limits to capture the singular structure in these regions of phase-space.
The dipole factorisation formula can be written schematically as
\begin{eqnarray}\label{eq:factorisation}
    |\mathcal{M}_{n+1}|^{2} \rightarrow |\mathcal{M}_{n}|^{2} \otimes \mathbf{V}_{ij,k} \, ,
\end{eqnarray}
where $\mathbf{V}_{ij,k}$ is a process independent, singular factor. It depends on the momenta and quantum numbers (colour and spin) of partons $i,j,k$, where $i$ is the emitter parton, 
$j$ is the emitted parton, and $k$ is the spectator parton.
For singly unresolved limits, this factorisation isolates all the divergent behaviour in $\mathbf{V}_{ij,k}$ and the factor $|\mathcal{M}_{n}|^{2}$ is free of divergences, which makes 
it more amenable to emulation through a neural network.
The dipole factorisation formula forms the basis of our fitting ansatz which we present in detail in Section~\ref{sec:ansatz}.

\subsection{Fitting coefficients of Catani-Seymour dipoles}\label{sec:ansatz}
Instead of using a neural network to fit the matrix element directly, we use the dipole factorisation formula to build an ansatz of the colour and helicity summed $n+1$-body matrix element,
\begin{eqnarray}\label{eq:ansatz}
    \ME{n+1} = \sum_{\{ijk\}} C_{ijk} D_{ij,k} \, ,
\end{eqnarray}
where $D_{ij,k} = \langle V_{ij,k} \rangle / s_{ij}$ are the spin-averaged Catani-Seymour dipoles divided by the corresponding Mandelstam invariant and $C_{ijk}$ are the coefficients we train the neural network to fit.
$C_{ijk}$ can be interpreted as the reduced matrix element in $n$-body phase-space. 
Since the input for the $C_{ijk}$ function is the full $n+1$ phase-phase information, the neural network will also model the phase-space mappings usually introduced in the factorisation formula.
A schematic diagram illustrating our ansatz is given in Figure~\ref{fig:nn_architecture}.
The sum over ${\{ijk\}}$ denotes the sum over relevant permutations of the external outgoing legs.
More detail on this is given in Section~\ref{sec:architecture}.
The representation (\ref{eq:ansatz}) is not unique but 
through appropriate training, the neural network takes advantage of the right ingredients to model 
the divergent soft and collinear behaviour of the matrix elements.

@@

\subsection{Data generation}\label{sec:data}
For all multiplicities, phase-space is sampled uniformly using the RAMBO algorithm \cite{Kleiss:1985gy} with a centre-of-mass energy $\sqrt{s_{\mathrm{com}}} = 1000$ GeV.
Phase-space points are subsequently clustered using {\FastJet} \cite{Cacciari:2011ma,noel_dawe_2021_4446849} with the $e^{+}e^{-}$ $k_{t}$ algorithm \cite{Catani:1991hj}.
Global phase-space cuts are applied according to the criterion $y_{\mathrm{cut}} \leq y_{ij}$ where $y_{ij}$ are the Mandelstam invariants normalised by $s_{\mathrm{com}}$.
Jets are clustered exclusively where $d_{\mathrm{cut}}$ was supplied to {\FastJet}. We took $d_{\mathrm{cut}} = \mathrm{max}(2 \times y_{\mathrm{cut}}, 0.01 \times s_{\mathrm{com}})$.
We explore three different values of the global phase-space cut parameter, $y_{\mathrm{cut}} = [0.01, 0.001, 0.0001]$, to demonstrate the ability of the factorisation-aware neural network to effectively interpolate in more and more singular regions of phase-space.

@@

Along with the 4-momenta, we also include the recoil factors
\begin{eqnarray}\label{eq:recoil_factor}
    y_{ij,k} = \dfrac{p_{i}p_{j}}{p_{i}p_{j} + p_{j}p_{k} + p_{i}p_{k}} \, ,
\end{eqnarray}
as input for relevant permutations of $\{ijk\}$.
We take the natural logarithm of $y_{ij,k}$ before standardising them to a zero mean and unit variance across the training dataset.
We find that the addition of these recoil factors significantly improves performance of our neural network emulator during training and testing.
This can be attributed to the fact that the coefficients $C_{ijk}$ rely heavily on the mapped momenta in $n$-body phase-space, where $y_{ij,k}$ is usually required when performing the momentum mapping, see e.g. \cite{Catani:1996vz}.

Following our ansatz (\ref{eq:ansatz}), we must provide spin-averaged Catani-Seymour dipoles to the neural network. These dipoles are computed for all phase-space points before training begins, but they are not passed directly to the neural network as input features. Instead, they are only included in our custom loss function which will be explained in the next subsection.

@@

\paragraph*{Training, validation, and testing datasets}
In Ref. \cite{Badger:2020uow}, the authors used 500k training samples to train their models.
For a fair comparison of our respective methods, we follow their methodology closely by constructing our models to be as close as possible to theirs and generate training and testing data by using code from their project repository \cite{n3jet_repo}\footnote{For our main results we generate all data using our own code}.
Our model architectures will not be identical due to the difference in our methods, details of our model architecture are given in Section~\ref{sec:architecture}.
Results of this comparison are presented in Section~\ref{sec:n3jet_comparison}.

@@

These choices of hyperparameters were not results of extensive scanning of parameter space and were chosen heuristically.
Hyperparameters can be tuned more optimally using more sophisticated methods, but they all rely on training a large number of models which is computationally expensive and time-consuming.

\paragraph*{Custom loss function}
To assess the model's performance when training we need to compare the network predictions with the targets. Our metric for the model's regression performance is the mean squared error (MSE)
\begin{eqnarray}\label{eq:mse}
    L_{\mathrm{MSE}} = \dfrac{1}{N} \sum_{i=1}^{N} \left(y_{i} - p(\vec{x}_{i}; \theta)\right)^{2} \, ,
\end{eqnarray}
where $y_{i}$ is the target for the $i$-th sample out of $N$ samples, and $p(\vec{x}_{i}; \theta)$ is the corresponding prediction
obtained by combining the dipole factors and the azimuthal dependency terms multiplied by their NN-learned coefficients. 
In order to correct for the scale difference, we need to apply the transformation (\ref{eq:arcsinh}) on the neural network prediction first, with the same $S_{\mathrm{pred}}$ that was used for scaling the targets.
We choose the mean squared error to measure the network's performance because it is sensitive to outliers in the target distribution.
This is useful because even though we have taken measures to reduce the span of the targets, the distribution still contains a tail towards larger values which correspond to soft and collinear configurations.
These points have large contributions to the cross-section when integrating over phase-space, meaning it is important that we accurately predict these points.
It is also convenient that the mean squared error tends to learn the mean of the target distribution which, in our case, corresponds to the cross-section.

@@

The form of $L_{\mathrm{pen}}$ is reminiscent of the usual L1 regularisation which promotes sparse models.
Regularisation is usually included to prevent overfitting by making the model make decisions on the most important features, reducing other features to zero.
In our case we would like the neural network to learn about the universal factorisation property in QCD by making it choose a minimal amount of dipoles to represent the matrix element in singular regions.
In addition to preventing overfitting, (\ref{eq:penalty}) helps the neural network to extrapolate to more soft and collinear regions as it has learnt to choose which dipoles are relevant in specific configurations.
For non-singular configurations, the neural network is free to interpolate as there is not a clear set of dipoles that dominate, meaning $L_{\mathrm{pen}}$ is small.

@@

\subsection{Comparison with previous work}\label{sec:n3jet_comparison}
We compare our method to methods for tree-level matrix elements emulation from Ref.~\cite{Badger:2020uow}.
The authors presented two methods: `single' and `ensemble' models. A `single' model indicates that there is 
one neural network trained across the entire phase-space, and an `ensemble' model indicates that there is a group of neural networks trained together
with weighting functions that focus individual networks on a specific divergent region of phase-space.
In order to conduct a fair comparison, we have made attempts to follow their training methodology closely by 
constructing the neural networks at the centre of our model with a similar structure (e.g. same hidden layer structure, same activation functions, 
same random initial weight distribution), training methodology (same {\EarlyStopping} criteria, same initial learning rate), and have generated training and testing datasets using code 
available on the authors' project repository with the relevant cuts.

We compare the distribution of errors for the matrix elements predictions on 3 million phase-space points.
The distribution of errors is crucial because it informs us of the performance of our emulators at the matrix element level.
It is well known that a neural network optimised with the mean squared error loss function has tendencies to learn the mean of the target distribution \cite{Nachman_2020},
meaning the quality of the cross-section prediction can potentially belie the point-to-point accuracy of emulators.
In Figure \ref{fig:comparison}, we plot the distribution of errors for matrix element predictions where our method is labelled `Dipole NN'.
We compare against the `single' and `ensemble' methods by training and testing using the corresponding datasets.
For more detailed information on the differences please refer to \cite{Badger:2020uow}.
Note that the height of the peak for the dipole histograms are not illustrated in the figure as it would not fit on the current axes but that is not important for this comparison.
We can see that the prediction-to-truth ratio distribution for our method is much narrower and consistently peaked around the ideal accuracy, indicating our model performs better on a per-point basis for all multiplicities.
Even with this reduced NN size we can see that incorporating the known
divergent structure explicitly in the model gives better results, as it
uses the NN representation to learn a function that is more
suitably approximated by a NN.
For example, even though the three jet matrix element has a fairly trivial analytical structure, a standard
fitting approach using a NN typically struggles to reproduce
divergences. In our approach the NN only needs to emulate a non-singular
modulation on top of the main divergent behaviour and is therefore more
suited to the task.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{fame1/n3jet_comparison_inset.pdf}
    \caption{Error distribution compared to Figure 3 in Ref.~\cite{Badger:2020uow}, where data to reproduce the histograms were provided by the authors.
    We plot the log ratio of the matrix element as predicted by the neural network ensemble and the value from {\NJet} on the main axes for comparison. The blue and orange dipole histograms representing our method 
    are cut off at the top on the main axes, but the most important feature is the narrowness of the peak centred around the ideal value.
    The insets show the detailed distribution of our result on a linear scale.}
    \label{fig:comparison}
\end{figure}

@@

By increasing the size of the training datasets we aim to expose the neural network to more samples of the phase-space, thereby increasing accuracy on predictions on as much of the phase-space as possible.
Along with increasing the number of training samples, the neural network architecture has been expanded to include more hidden nodes and hidden layers.
The extra hidden nodes and layers introduces more parameters into the model allowing the neural network to utilise the additional data. 
We found that to get good performance, we had to balance the size of the training dataset used and the size of the network. i.e. a small network is not expected to capture all the variations in a large dataset as easily as a larger network would, due to the smaller number of parameters available to the network.
Of course, we also had to consider more physical constraints such as the time spent on training the neural networks which limits both the size of training datasets and architecture.
In Figure~\ref{fig:error_500k_vs_40m}, we show the improvements in accuracy of our main NNs compared to the smaller NNs from Section~\ref{sec:n3jet_comparison}.
Although the training and testing data for the smaller NNs are not identical due to differences in code used to generate the sets\footnote{Although RAMBO is used for phase-space generation in both works, the selection criteria is different (JADE algorithm vs $k_{t}$ algorithm).}, for this comparison it suffices to show that the larger NNs are orders of magnitude more accurate than the smaller NNs.

@@

NN predictions are depicted as coloured scatter plots where the colour indicates the value of the minimum $s_{ij}$ between any pair of final-state particles at that phase-space point.
To more easily visualise the extrapolation performance of the NNs we highlight the regions where the minimum $s_{ij}$ goes below the global phase-space cut applied to the training set the models were trained on, 
for each cut made. The regions of the plots where this occurs have been coloured in red.  
With these trajectories being completely randomly selected in phase-space there is a possibility for there to be doubly singular points or worse.
To check for this we used {\FastJet} to cluster the phase-space points in the same way we did for data generation, see Section~\ref{sec:data}.
The pink regions indicate points which have two separate single unresolved limits, we label this configuration `Single+single'.
The purple regions indicate points which have a double unresolved limit (i.e. three particles in one jet), we label this configuration `Double'.
The blue regions indicate points which have both a double unresolved limit and a separate single unresolved limit, we label this configuration `Double+single'.
We do not include the quark-anti-quark invariant in defining these regions as there is no associated infrared divergence.
The pink, purple, and blue bands indicates regions of points which would have been discarded for our training and testing datasets.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{fame1/trajectory_1271_double.pdf}
    \caption{Left: matrix element prediction of random phase-space trajectory.
    Right: ratio of NN and {\NJet}.
    NN predictions are coloured by minimum $s_{ij}$ pair.
    Red bands indicate min($s_{ij}$) is smaller than $y_{\mathrm{cut}}$.
    Pink bands indicate where there are two separate single unresolved limits.
    Purple bands indicate double unresolved limits, i.e. three particles in one jet.
    Blue bands indicate where there is both a double and a single unresolved limit.
    The pink, purple, and blue bands represent regions of phase-space which would have been excluded by {\FastJet}.}
    \label{fig:random_trajectory}
\end{figure}

>>>>

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Emulation of tree-level $e^{+}e^{-}$ to jets matrix elements}
\label{chapter:fame1}

\section{Motivation}
\label{sec:intro}
The capacity of neural networks to approximate intricate functions have already been 
used to provide fast calculations of production cross-sections~\cite{Otten:2018kum,Buckley:2020bxg}.
In this chapter we investigate whether NNs can approximate production cross-sections more differentially by replacing computationally expensive 
matrix element calculations. 
The challenge with this endeavour is that matrix elements are plagued with numerous divergences that arise from infrared divergences.
In previous works \cite{Badger:2020uow,Aylett-Bullock:2021hmo} a combination of individual neural networks were used 
to approximate matrix elements. In order to deal with the complex structure of the matrix elements 
the authors of these studies divided the phase-space into sectors according to the infrared singularities and trained networks on these sectors,
thereby limiting the complexity of the fit by isolating a single divergence per sector. All sectors were then combined to make the final prediction.
While the authors of the initial study for electron-positron annihilation showed good agreement between the total cross-section and histogrammed distributions both at LO and NLO,
we note that the accuracy of the interpolation at the level of individual points was a lot worse than when averaged in the 
histograms. In addition, the performance of the extrapolation outside of its training phase-space 
(i.e. more singular configurations than those considered to fit the model) is problematic.

@@

This chapter is organised as follows. 
Section~\ref{sec:fitting} introduces our factorisation-aware 
deep neural network model, Section~\ref{sec:results} showcases our results with a comparison with the 
model from Ref.~\cite{Badger:2020uow} in Section~\ref{sec:n3jet_comparison}, and an analysis of the performance of our model in Section~\ref{sec:mainresult}. 
To show that the neural network is both interpolating and extrapolating well we show its 
behaviour on random trajectories in phase-space in Section~\ref{sec:trajectories}. 
Finally, the findings of this chapter will be summarised in Section~\ref{sec:conclusion}.
Some discussion of specific details are collected in Appendix~\ref{appendix:fame1}
so as not to distract from the main discussion.

@@

A neural network can be seen as a function $f(\mathbf{x}; \theta) = \mathbf{y}$, where $f: \mathbb{R}^{d} \rightarrow \mathbb{R}$ maps a 
$d$-dimensional vector $\mathbf{x}$ of inputs onto the vector of outputs $\mathbf{y}$, and where $\theta$ are the parameters of the neural network which we aim to optimise 
such that the outputs $\mathbf{y}$ of the neural network match the target as well as possible. 
The simplest implementation of an emulator would be to take the input as the kinematic information of the phase-space point and the output to be the full matrix element. 
Our approach modifies both the input of the NN and its target, as described in the following sections.

\subsection{Infrared divergences and dipole factorisation formula}
It is well known that in soft and collinear limits the matrix element in $(n+1)$-body phase-space factorises into a singular factor and a reduced matrix element in $n$-body phase-space \cite{Altarelli:1977zs,BASSETTO1983201} (Section~\ref{sec:me_factorisation}).
This factorisation was used by Catani and Seymour \cite{Catani:1996vz} to construct subtraction terms for the real radiation part of an NLO calculation (Section~\ref{sec:CS_dipoles}).
They introduced a factorisation formula with universal dipoles that smoothly interpolates 
between the soft and collinear limits to capture the singular structure in these regions of phase-space.
The dipole factorisation formula can be written schematically as
\begin{eqnarray}\label{eq:factorisation}
    |\mathcal{M}_{n+1}|^{2} \rightarrow |\mathcal{M}_{n}|^{2} \otimes \mathbf{V}_{ij,k} \, ,
\end{eqnarray}
where $\mathbf{V}_{ij,k}$ is a process independent, singular factor. It depends on the momenta and quantum numbers (colour and spin) of partons $i,j,k$, where $i$ is the emitter parton, 
$j$ is the emitted parton, and $k$ is the spectator parton.
For singly unresolved limits, this factorisation isolates all the divergent behaviour in $\mathbf{V}_{ij,k}$ and the factor $|\mathcal{M}_{n}|^{2}$ is free of divergences, which makes 
it more amenable to emulation through a neural network.
The dipole factorisation formula forms the basis of our fitting ansatz which we present in detail in Section~\ref{sec:ansatz}.

\subsection{Fitting coefficients of Catani-Seymour dipoles}\label{sec:ansatz}
Instead of using a neural network to fit the matrix element directly, we use the dipole factorisation formula to build an ansatz of the colour and helicity summed $(n+1)$-body matrix element,
\begin{eqnarray}\label{eq:ansatz}
    \ME{n+1} = \sum_{\{ijk\}} C_{ijk} D_{ij,k} \, ,
\end{eqnarray}
where $D_{ij,k} = \langle V_{ij,k} \rangle / s_{ij}$ are the spin-averaged Catani-Seymour dipoles divided by the corresponding Mandelstam invariant and $C_{ijk}$ are the coefficients we train the neural network to fit.
$C_{ijk}$ can be interpreted as the reduced matrix element in $n$-body phase-space. 
Since the input for the $C_{ijk}$ function is the full $(n+1)$ phase-phase information, the neural network will also model the phase-space mappings usually introduced in the factorisation formula.
A schematic diagram illustrating our ansatz is given in Figure~\ref{fig:nn_architecture}.
The sum over ${\{ijk\}}$ denotes the sum over relevant permutations of the external outgoing legs.
More detail on this is given in Section~\ref{sec:architecture}.
The representation in Eq. (\ref{eq:ansatz}) is not unique but 
through appropriate training, the neural network takes advantage of the right ingredients to model 
the divergent soft and collinear behaviour of the matrix elements.

@@

\subsection{Data generation}\label{sec:data}
For all multiplicities, phase-space is sampled uniformly using the {\RAMBO} algorithm \cite{Kleiss:1985gy} with a centre-of-mass energy $\sqrt{s_{\mathrm{com}}} = 1000$ GeV.
Phase-space points are subsequently clustered using {\FastJet} \cite{Cacciari:2011ma,noel_dawe_2021_4446849} with the $e^{+}e^{-}$ $k_{t}$ algorithm \cite{Catani:1991hj}.
Global phase-space cuts are applied according to the criterion $y_{\mathrm{cut}} \leq y_{ij}$ where $y_{ij}$ are the Mandelstam invariants normalised by $s_{\mathrm{com}}$.
Jets are clustered exclusively where $d_{\mathrm{cut}}$ was supplied to {\FastJet}. We took $d_{\mathrm{cut}} = \mathrm{max}(2 \times y_{\mathrm{cut}}, 0.01 \times s_{\mathrm{com}})$.
We explore three different values of the global phase-space cut parameter, $y_{\mathrm{cut}} = [0.01, 0.001, 0.0001]$, to demonstrate the ability of the factorisation-aware neural network to effectively interpolate in more and more singular regions of phase-space.

@@

Along with the 4-momenta, we also include the recoil factors
\begin{eqnarray}\label{eq:recoil_factor}
    y_{ij,k} = \dfrac{p_{i}p_{j}}{p_{i}p_{j} + p_{j}p_{k} + p_{i}p_{k}} \, ,
\end{eqnarray}
as input for relevant permutations of $\{ijk\}$.
We take the natural logarithm of $y_{ij,k}$ before standardising them to a zero mean and unit variance across the training dataset.
We find that the addition of these recoil factors significantly improves performance of our neural network emulator during training and testing.
This can be attributed to the fact that the coefficients $C_{ijk}$ rely heavily on the mapped momenta in $n$-body phase-space, where $y_{ij,k}$ is usually required when performing the momentum mapping, see e.g. Ref.~\cite{Catani:1996vz}.

Following our ansatz in Eq.~(\ref{eq:ansatz}), we must provide spin-averaged Catani-Seymour dipoles to the neural network. These dipoles are computed for all phase-space points before training begins, but they are not passed directly to the neural network as input features. Instead, they are only included in our custom loss function which will be explained in the next subsection.

@@

\paragraph*{Training, validation, and testing datasets}
In Ref. \cite{Badger:2020uow}, the authors used 500k training samples to train their models.
For a fair comparison of our respective methods, we follow their methodology closely by constructing our models to be as close as possible to theirs and generate training data by using code from their project repository \cite{n3jet_repo}\footnote{For our main results we generate all data using our own code}.
Our model architectures will not be identical due to the difference in our methods, details of our model architecture are given in Section~\ref{sec:architecture}.
Results of this comparison are presented in Section~\ref{sec:n3jet_comparison}.

@@

These choices of hyperparameters were not results of extensive scanning of parameter space and were chosen heuristically.
Hyperparameters can be tuned more optimally using more sophisticated methods as described in Section~\ref{sec:hyperparameters}, but they all rely on training a large number of models which is computationally expensive and time-consuming.

\paragraph*{Custom loss function}
To assess the model's performance when training we need to compare the network predictions with the targets. Our metric for the model's regression performance is the mean squared error (MSE)
\begin{eqnarray}\label{eq:mse}
    L_{\mathrm{MSE}} = \dfrac{1}{N} \sum_{i=1}^{N} \left(y_{i} - p(\mathbf{x}_{i}; \theta)\right)^{2} \, ,
\end{eqnarray}
where $y_{i}$ is the target for the $i$-th sample out of $N$ samples, and $p(\mathbf{x}_{i}; \theta)$ is the corresponding prediction
obtained by combining the dipole factors and the azimuthal dependency terms multiplied by their NN-learned coefficients. 
In order to correct for the scale difference, we need to apply the transformation from Eq.~(\ref{eq:arcsinh}) on the neural network prediction first, with the same $S_{\mathrm{pred}}$ that was used for scaling the targets.
We choose the mean squared error to measure the network's performance because it is sensitive to outliers in the target distribution.
This is useful because even though we have taken measures to reduce the span of the targets, the distribution still contains a tail towards larger values which correspond to soft and collinear configurations.
These points have large contributions to the cross-section when integrating over phase-space, meaning it is important that we accurately predict these points.
It is also convenient that the mean squared error tends to learn the mean of the target distribution which, in our case, corresponds to the cross-section.

@@

The form of $L_{\mathrm{pen}}$ is reminiscent of the usual L1 regularisation which promotes sparse models.
Regularisation is usually included to prevent overfitting by making the model make decisions on the most important features, reducing other features to zero.
In our case we would like the neural network to learn about the universal factorisation property in QCD by making it choose a minimal amount of dipoles to represent the matrix element in singular regions.
In addition to preventing overfitting, Eq.~(\ref{eq:penalty}) helps the neural network to extrapolate to more soft and collinear regions as it has learnt to choose which dipoles are relevant in specific configurations.
For non-singular configurations, the neural network is free to interpolate as there is not a clear set of dipoles that dominate, meaning $L_{\mathrm{pen}}$ is small.

@@

\subsection{Comparison with previous work}\label{sec:n3jet_comparison}
We compare our method to methods for tree-level matrix elements emulation from Ref.~\cite{Badger:2020uow}.
The authors presented two methods: `single' and `ensemble' models. A `single' model indicates that there is 
one neural network trained across the entire phase-space, and an `ensemble' model indicates that there is a group of neural networks trained together
with weighting functions that focus individual networks on a specific divergent region of phase-space.
In order to conduct a fair comparison, we have made attempts to follow their training methodology closely by 
constructing the neural networks at the centre of our model with a similar structure (e.g. same hidden layer structure, same activation functions, 
same random initial weight distribution), training methodology (same {\EarlyStopping} criteria, same initial learning rate), and have generated training datasets using code 
available on the authors' project repository with the relevant cuts.
For the testing data, we thank the authors for providing the same testing set used in their publication.

We compare the distribution of errors for the matrix elements predictions on 3 million phase-space points.
The distribution of errors is crucial because it informs us of the performance of our emulators at the matrix element level.
It is well known that a neural network optimised with the mean squared error loss function has tendencies to learn the mean of the target distribution \cite{Nachman_2020},
meaning the quality of the cross-section prediction can potentially belie the point-to-point accuracy of emulators.
In Figure \ref{fig:comparison}, we plot the distribution of errors for matrix element predictions where our method is labelled `Dipole NN'.
We compare against the `single' and `ensemble' methods by training and testing using the corresponding datasets.
Note that the height of the peak for the dipole histograms are not illustrated in the figure as it would not fit on the current axes but that is not important for this comparison.
We can see that the prediction-to-truth ratio distribution for our method is much narrower and consistently peaked around the ideal accuracy, indicating our model performs better on a per-point basis for all multiplicities.
Even with this reduced NN size we can see that incorporating the known
divergent structure explicitly in the model gives better results, as it
uses the NN representation to learn a function that is more
suitably approximated by a NN.
For example, even though the three jet matrix element has a fairly trivial analytical structure, a standard
fitting approach using a NN typically struggles to reproduce
divergences. In our approach the NN only needs to emulate a non-singular
modulation on top of the main divergent behaviour and is therefore more
suited to the task.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{fame1/n3jet_comparison_inset.pdf}
    \caption{Error distribution compared to Figure 3 in Ref.~\cite{Badger:2020uow}, where data to reproduce the histograms were provided by the authors.
    We plot the log ratio of the matrix element as predicted by the neural network ensemble and the value from {\NJet} on the main axes for comparison. The blue and orange dipole histograms representing our method 
    are cut off at the top on the main axes, but the most important feature is the narrowness of the peak centred around the ideal value of 0.
    The insets show the detailed distribution of our result on a linear scale.}
    \label{fig:comparison}
\end{figure}

@@

By increasing the size of the training datasets we aim to expose the neural network to more samples of the phase-space, thereby increasing accuracy on predictions on as much of the phase-space as possible.
Along with increasing the number of training samples, the neural network architecture has been expanded to include more hidden nodes and hidden layers.
The extra hidden nodes and layers introduces more parameters into the model allowing the neural network to utilise the additional data. 
We found that to get good performance, we had to balance the size of the training dataset used and the size of the network. i.e. a small network is not expected to capture all the variations in a large dataset as easily as a larger network would, due to the smaller number of parameters available to the network.
Of course, we also had to consider more physical constraints such as the time spent on training the neural networks which limits both the size of training datasets and architecture.
In Figure~\ref{fig:error_500k_vs_40m}, we show the improvements in accuracy of our main NNs compared to the smaller NNs from Section~\ref{sec:n3jet_comparison}.
Although the training and testing data for the smaller NNs are not identical due to differences in code used to generate the sets\footnote{Although {\RAMBO} is used for phase-space generation in both works, the selection criteria is different (JADE algorithm vs $k_{t}$ algorithm).}, for this comparison it suffices to show that the larger NNs are orders of magnitude more accurate than the smaller NNs.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{fame1/trajectory_1271_double.pdf}
    \caption{Left: matrix element prediction of random phase-space trajectory.
    Right: ratio of NN and {\NJet}.
    NN predictions are coloured by minimum $s_{ij}$ pair.
    Red bands indicate min($s_{ij}$) is smaller than $y_{\mathrm{cut}}$.
    Pink bands indicate where there are two separate single unresolved limits.
    Purple bands indicate double unresolved limits, i.e. three particles in one jet.
    % Blue bands indicate where there is both a double and a single unresolved limit.
    The pink, and purple bands represent regions of phase-space which would have been excluded by {\FastJet}.}
    \label{fig:random_trajectory}
\end{figure}

NN predictions are depicted as coloured scatter plots where the colour indicates the value of the minimum $s_{ij}$ between any pair of final-state particles at that phase-space point.
To more easily visualise the extrapolation performance of the NNs we highlight the regions where the minimum $s_{ij}$ goes below the global phase-space cut applied to the training set the models were trained on, 
for each cut made. The regions of the plots where this occurs have been coloured in red.  
With these trajectories being completely randomly selected in phase-space there is a possibility for there to be doubly singular points or worse.
To check for this we used {\FastJet} to cluster the phase-space points in the same way we did for data generation, see Section~\ref{sec:data}.
The pink regions indicate points which have two separate single unresolved limits, we label this configuration `Single+single'.
The purple regions indicate points which have a double unresolved limit (i.e. three particles in one jet), we label this configuration `Double'.
Although not seen in Figure~\ref{fig:random_trajectory}, there are points which have both a double unresolved limit and a separate single unresolved limit,
we label this configuration `Double+single' and indicate is as a blue region\footnote{Another random phase-space trajectory is presented in Appendix~\ref{appendix:fame1_trajectory_generation} which contains these `double+single' regions.}.
We do not include the quark-anti-quark invariant in defining these regions as there is no associated infrared divergence.
The pink, purple, and blue bands indicates regions of points which would have been discarded for our training and testing datasets.

@@

\begin{table}
    \centering
    \small
    \begin{tabular}{|c|c|c|c|c|c|}
        \hline
        Region & $y_{\mathrm{cut}}$ & Frac. of pts & Pts outside $0.1\%$ & Pts outside $1\%$ & Pts outside $5\%$ \\
        \hline
        \multirow{3}{*}{White} & 0.01   & 36.5\% & 1.9\% & 0.0\% & 0.0\% \\
        \cline{2-6}
                               & 0.001  & 74.8\% & 0.98\% & 0.0\% & 0.0\% \\
        \cline{2-6}
                               & 0.0001 & 78.4\% & 1.9\% & 0.0\%  & 0.0\% \\
        \hline
        \hline
        \multirow{3}{*}{Pink}  & 0.01   & 27.6\% & 58.2\% & 25.2\% & 12.9\% \\
        \cline{2-6}
                               & 0.001  & 13.5\% & 31.7\% & 6.5\% & 0.53\% \\
        \cline{2-6}
                               & 0.0001 & 13.5\% & 38.9\% & 9.3\% & 2.8\% \\
        \hline
        \hline
        \multirow{3}{*}{Purple}& 0.01   & 7.2\% & 69.1\% & 32.1\% & 20.3\% \\
        \cline{2-6}
                               & 0.001  & 3.6\% & 31.3\% & 7.9\% & 4.5\% \\
        \cline{2-6}
                               & 0.0001 & 3.6\% & 30.5\% & 1.6\% & 0.0\% \\
        \hline
        \hline
        \multirow{3}{*}{Blue}  & 0.01   & 1.5\% & 79.3\% & 41.6\% & 0.77\% \\
        \cline{2-6}
                               & 0.001  & 0.4\% & 29.2\% & 0.0\% & 0.0\% \\
        \cline{2-6}
                               & 0.0001 & 0.4\% & 76.4\% & 0.0\% & 0.0\% \\
        \hline
        \hline
        \multirow{3}{*}{Red}   & 0.01   & 52.5\% & 76.2\% & 38.6\% & 20.5\% \\
        \cline{2-6}
                               & 0.001  & 7.0\%  & 69.9\% & 25.2\% & 9.5\% \\
        \cline{2-6}
                               & 0.0001 & 1.1\%  & 90.8\% & 33.8\% & 1.5\% \\
        \hline
    \end{tabular}
    \caption{The performance of trajectory predictions separated for white, pink, purple, blue, and red regions.
    We present the percentage of points that lie outside 0.1\%/1.0\%/5.0\% errors.
    Fraction of points indicates the percentage of points that lie in the region of interest, out of all phase-space points from the 50 random trajectories we examined.
    }
    \label{table:trajectory_predictions}
\end{table}
